Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2020-02-09 paper.Rnw
edits to sec 5

Our intention is to develop, demonstrate, and contrast these three
approaches to modeling log data.
A rich, and often polemic, literature surrounds principal
stratification and mediation analysis
\citep[e.g.][]{rubin2004direct,vanderweele2011principal,pearl2011principal,mealli2012refreshing,vanderweele2012comments}.
\citet{vanderweele2008simple}, \citet{jo2008causal} and others have
derived formal equivalencies between parameters under certain
conditions.
Our contribution will build on this literature in two ways: first, by
providing a detailed demonstration and comparison of matching, mediation analysis,
and principal stratification on the same dataset, and second by
focusing on the applicability of these three methods in the specific
context of studying implementation data from an randomized
experiment studying educational technology (EdTech).

@@

Dichotomizing hint requests into $H$, instead of $\bar{h}$, re-defines
the causal question.
To make sense of the respective roles of $H$ and $\bar{h}$, the
continuous measure on which it is based (see Section
\ref{sec:dichotomizingHintUsage}),
we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\bar{h}$.
In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
in the treatment group \Sexpr{round((1-mixProp)*100)}\% of students have $H=1$ and \Sexpr{round((mixProp)*100)}\% have
$H=0$, and $\bar{h}$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

@@

The goal of principal stratification is to estimate ``principal
effects'':
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\bar{h}(1)=r]
\end{equation*}
This is the treatment effect for subjects who \emph{would} request
hints at level $\bar{h}(1)=r$, if assigned to the treatment condition.
Whereas the focus of the observational study and mediation
analysis was the relationship between $\bar{h}$ and posttest scores, the
focus of PS is the relationship between $\bar{h}$ and treatment effects.
Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\bar{h}(1)=r]$,
despite the fact that $\bar{h}(1)$ is never observed at the same time as
$Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
strictly counterfactual quantities, as in mediation analysis.
$\bar{h}(1)$ is defined for subjects in both randomized groups, but only
observed in the treatment group.

In previous sections $\bar{h}$ was dichotomized into $H$; such
dichotomization is not necessary here.
However, $\bar{h}$ has some disadvantages as a continuous measure of
hint usage.
For one, since it is essentially a sample mean over a subset of each student's
worked problems, much of its variance is driven by the total number of
problems students worked---in particular, its extreme values belong to
those students who barely used the tutor at all.
Further, it does not account for varying difficulty of the tutor's
problems.
In Section \ref{sec:dichotomizingHintUsage}, we showed that
dichotomized $\bar{h}$ largely agrees with a dichotomized version of a
more sophisticated measure of hint usage.
However, when modeling hint usage continuously, the correspondence may
not hold \citet[see][for a more complete discussion]{aoas}.

For those reasons, we modeled hint usage at the problem level, with
equation \eqref{eq:rasch}, but with two important differences.
First, a conceptual difference: the $\eta$ in \eqref{eq:rasch} was
replaced by $\eta (1)$, indicating that it is measuring
\emph{potential} hint usage---the hint usage a student would exhibit
were he or she assigned to the treatment condition.
Whereas $\eta$ is only defined for students assigned to the treatment condition,
$\eta(1)$ is defined for all the students in the study.
Second, a modeling difference: since we are not dichotomizing $\eta(1)$
in this analysis, we did not model it as a normal mixture, as in
\eqref{eq:mixture}.

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the principal stratification model, replacing $\eta(1)$ for
$\bar{h}(1)$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

\subsection{Identification and Estimation}

@@

Figure \ref{fig:psResults} shows the estimated linear function
$\hat{\tau}{\eta(0)}$.
The left panel shows the CTA1 treatment effect, $Y(1)-Y(0)$ as
$\tau(1)$ varies; the black line shows the estimate (i.e. posterior
mean) and the red lines are random draws from the posterior
distribution, showing estimation uncertainty.
Eighty-eight percent of posterior draws of the slope of the
$\hat{\tau}(\cdot)$ function were positive, implying a posterior
probability of 0.88 that students who asked for hints
with greater regularity benefited more from the CTA1 curriculum.
The right hand panel plots one posterior draw of $\eta(1)$
alongside observed outcomes $Y$.
The figure also includes estimated regression lines for the two
treatment groups.
Although, as observed in Sections \ref{sec:observational} and
\ref{sec:mediation}, $\eta(1)$ is anticorrelated with $Y$, the slope
is steeper in the control group than in the treatment group.
The distance between the two regression lines is the treatment effect,
which grows with $\eta(1)$.

@@

@@

In contrast, no ignorabilty assumptions are necessary for principal
stratification.
% That said, an assumption similar to ignorability may play an important
% role for interpretation.
However, principal stratification estimators require modeling
assumptions---most problematically, $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ of
\eqref{eq:posteriorEta}, the model for $Y(0)$
conditional on $\eta(1)$, $\bm{x}$, and parameters $\bm{\theta}$.
While any model specification can, and should, be checked against
observed data, this process can only narrow space of acceptable
models, it cannot be used to determine the correct form of
$p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$.
This is because $Y(0)$ and $\eta(1)$ are never observed simultaneously.
Therefore, it is possible for an analyst to misspecify $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$, such that principal effect estimates are
severely biased, without being able to detect the model
misspecification with the data.
That is, the data will typically be unable to distinguish between
alternative $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ models that lead to
qualitatively different conclusions.
In practice, analysts make the untestable assumption that $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ is
of the same form as $p(Y|\bm{x},Z=1,\bm{\theta},\eta(1))$.

>>>>

Our intention is to develop, demonstrate, and contrast these three
approaches to modeling log data.
A rich, and often polemic, literature surrounds principal
stratification and mediation analysis
\citep[e.g.][]{rubin2004direct,vanderweele2011principal,pearl2011principal,mealli2012refreshing,vanderweele2012comments}.
\citet{vanderweele2008simple}, \citet{jo2008causal} and others have
derived formal equivalences between parameters under certain
conditions.
Our contribution will build on this literature in two ways: first, by
providing a detailed demonstration and comparison of matching, mediation analysis,
and principal stratification on the same dataset, and second by
focusing on the applicability of these three methods in the specific
context of studying implementation data from an randomized
experiment studying educational technology (EdTech).

@@

Dichotomizing hint requests into $H$, instead of $\bar{h}$, re-defines
the causal question.
To make sense of the respective roles of $H$ and $\bar{h}$, the
continuous measure on which it is based (see Section
\ref{sec:dichotomizingHintUsage}),
we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\bar{h}$.
In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
in the treatment group \Sexpr{round((1-mixProp)*100)}\% of students have $H=1$ and \Sexpr{round((mixProp)*100)}\% have
$H=0$, and $\bar{h}$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely under-sells the role
of hints, but allows a tractable mediation analysis.

@@

The goal of principal stratification is to estimate ``principal
effects'':
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\bar{h}(1)=r]
\end{equation*}
This is the treatment effect for subjects who \emph{would} request
hints at level $\bar{h}(1)=r$, if assigned to the treatment condition.
Whereas the goal of the observational study and mediation
analysis was the relationship between $\bar{h}$ and posttest scores, the
goal of PS is the relationship between $\bar{h}(1)$ and treatment effects.
Notably, although $Y(0)$ and $\bar{h}(1)$ and $Y(0)$ are never
simultaneously observed, the $\tau(r)$ estimand does not depend on any
strictly-counterfactual outcomes.
Unlike, say, $Y(0,H(1))$, $Y(0)$ and $Y(1)$ are each truly
potential---they would each occur under treatment assignments $Z=0$
and $Z=1$.
That said, estimating $\tau(r)$ requires estimation of
$\EE[Y(0)|\bar{h}(1)]$, the average of $Y(0)$ over a subset of the
sample that is unobserved and must be inferred.
For this reason, $\tau(r)$ is only \emph{partially} identified
\citep[e.g.][]{mealli2016identification}; even with an infinite
sample, principal effect estimates will still contain uncertainty.

In previous sections $\bar{h}$ was dichotomized into $H$; such
dichotomization is not necessary here.
However, $\bar{h}$ has some disadvantages as a continuous measure of
hint usage.
For one, since it is essentially a sample mean over a subset of each student's
worked problems, much of its variance is driven by the total number of
problems students worked---in particular, its extreme values belong to
those students who barely used the tutor at all.
Further, it does not account for varying difficulty of the tutor's
problems.
In Section \ref{sec:dichotomizingHintUsage}, we showed that
dichotomized $\bar{h}$ largely agrees with a dichotomized version of a
more sophisticated measure of hint usage.
However, when modeling hint usage continuously, the correspondence may
not hold \citep[see][for a more complete discussion]{aoas}.

For those reasons, we modeled hint usage at the problem level, with
equation \eqref{eq:rasch}, but with two important differences.
First, a conceptual difference: the $\eta$ in \eqref{eq:rasch} was
replaced by $\eta (1)$, indicating that it is measuring
\emph{potential} hint usage---the hint usage a student would exhibit
were he or she assigned to the treatment condition.
Whereas $\eta$ is only defined for students assigned to the treatment condition,
$\eta(1)$ is defined for all the students in the study.
Second, a modeling difference: since we did not dichotomize $\eta(1)$
in this analysis, we modeled it as normal , not as a normal mixture, as in
\eqref{eq:mixture}.
With this change, the principal stratification estimand became
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
the treatment effect for students whose potential hint usage, if assigned to the treatment condition $\eta(1)=r$.
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

\subsection{Identification and Estimation}
Estimating $\tau(r)$ required estimating $\EE[Y(0)|\eta(1)=r]$,
despite the fact that hint requests never observed at the same time as
$Y(0)$ (in fact $\eta(1)$ isn't directly observed in either the
treatment or control group).
However, hint requests are observed, and the conditional distribution
$\eta(1)|\bm{x}$ is identified, among those students with $Z=1$;
because of randomization of $Z$, these inferences extend to the
control group as well.

@@

@@

Figure \ref{fig:psResults} shows the estimated linear function
$\hat{\tau}(r)$.
The left panel shows the CTA1 treatment effect, $Y(1)-Y(0)$ as
$r=\tau(1)$ varies; the black line shows the estimate (i.e. posterior
mean) and the red lines are random draws from the posterior
distribution, showing estimation uncertainty.
Eighty-eight percent of posterior draws of the slope of the
$\hat{\tau}(\cdot)$ function were positive, implying a posterior
probability of 0.88 that students who asked for hints
with greater regularity benefited more from the CTA1 curriculum.
The right hand panel plots one posterior draw of $\eta(1)$
alongside observed outcomes $Y$.
The figure also includes estimated regression lines for the two
treatment groups.
Although, as observed in Sections \ref{sec:observational} and
\ref{sec:mediation}, $\eta(1)$ is anticorrelated with $Y$, the slope
is steeper in the control group than in the treatment group.
The distance between the two regression lines is the treatment effect,
which grows with $\eta(1)$.

@@

These results suggest that students who, if assigned to the treatment
condition, would be more likely to request hints on challenging
problems, tend to experience higher treatment effects.
That is, it suggests that frequent hint requesters benefit more from
assignment to the CTA1 condition than infrequent hint requesters.

@@

In contrast, no ignorability assumptions are necessary for principal
stratification.
% That said, an assumption similar to ignorability may play an important
% role for interpretation.
However, principal stratification estimators require modeling
assumptions---most problematically, $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ of
\eqref{eq:posteriorEta}, the model for $Y(0)$
conditional on $\eta(1)$, $\bm{x}$, and parameters $\bm{\theta}$.
While any model specification can, and should, be checked against
observed data, this process can only narrow space of acceptable
models, it cannot be used to determine the correct form of
$p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$.
This is because $Y(0)$ and $\eta(1)$ are never observed simultaneously.
Therefore, it is possible for an analyst to misspecify $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$, such that principal effect estimates are
severely biased, without being able to detect the model
misspecification with the data.
That is, the data will typically be unable to distinguish between
alternative $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ models that lead to
qualitatively different conclusions.
In practice, analysts make the untestable assumption that $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ is
of the same form as $p(Y|\bm{x},Z=1,\bm{\theta},\eta(1))$.
